Antigenic phenotypes and phylogenetic metrics improve forecasts of seasonal influenza A/H3N2 evolution

### 2019-12-13 manuscript/main.tex
Rename "non-epitope mutations" metric to "mutational load"
Updates figures and main text.

Closes #64.

We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
All of these metrics except for HI cross-immunity and DMS mutational effects rely only on HA sequences.
The cross-immunity metrics estimate how antigenically distinct each sample at time $t$ is from previously circulating samples based on either genetic distance at epitope sites or $\log_{2}$ titer distance from HI measurements.
Increased antigenic drift relative to previously circulating samples is expected to correspond to increased viral fitness.
Non-epitope mutations are the simplest of the functional constraint metrics, measuring the number of putatively deleterious mutations that have accumulated in each sample since their ancestor in the previous season.
DMS mutational effects provide a more comprehensive biophysical model of functional constraint by measuring the beneficial or deleterious effect of each possible single amino acid mutation in HA from the background of a previous vaccine strain, A/Perth/16/2009.
The growth metrics estimate how successful populations of samples have been in the last six months based on either rapid branching in the phylogeny (LBI) or the change in clade frequencies over time (delta frequency).

@@

We hypothesized that fitness metrics associated with viral success such as epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while metrics associated with fitness penalties, like non-epitope mutations, would receive negative coefficients.
We reasoned that both LBI and delta frequency would individually outperform the mechanistic metrics as both of these growth metrics estimate recent clade success regardless of the mechanistic basis for that success.
Correspondingly, we expected that a composite model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics, as this model would include both primary fitness constraints acting on our simulated populations.

@@

As expected, the true fitness model outperformed all other models, estimating a future population within 6.82 $\pm$ 1.52 amino acids (AAs) of the observed future and surpassing the naive model in 32 of 33 (97\%) timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Table~\ref{table_simulated_model_selection}).
Importantly, the true fitness model accurately approximated the diversity of future populations coming within -0.29 $\pm$ 2.01 AAs of observed future diversity.
In contrast to the true fitness model, the naive model was an average of 8.97 $\pm$ 1.35 AAs from the future and an average of 2.26 $\pm$ 1.75 AAs from the future's observed diversity.
With the exception of epitope cross-immunity, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
LBI was the best of these models, with a distance to the future of 7.57 $\pm$ 1.85 AAs.
Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
The non-epitope mutations metric received a consistently negative coefficient with an average distance of 8.27 $\pm$ 1.35 AAs.

@@

Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric when effects of purifying selection were also represented in the model.
The cross-immunity fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \cite{Luksza:2014hj} that are not explicitly present in our simulations.
To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each sample's ancestor in the previous season.
This linear fitness metric did not outperform the cross-immunity metric (Table~\ref{table_simulated_model_selection}).
However, when we fit a composite model of the epitope ancestor and non-epitope mutations metrics, this model outperformed both the individual non-epitope mutations model and the composite cross-immunity and non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.

We hypothesized that a composite model of mutually beneficial metrics could better approximate the true fitness of simulated viruses than models based on individual metrics.
To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: non-epitope mutations and LBI.
This composite model outperformed all individual metrics with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
The coefficients for LBI and non-epitope mutations remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
These results support our hypothesis that multiple complementary metrics can produce more accurate models.

@@

Finally, we tested all of our models on out-of-sample data.
Specifically, we fixed the coefficients of each model to the average values across the training/validation period and applied the resulting models to the next 9 years of previously unobserved simulated data.
A standard expectation from machine learning is that models will perform worse on test data than on validation data.
Instead, we found that nearly all models except for the true fitness and delta frequency performed as well or better with out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
The composite model of non-epitope mutations and LBI outperformed the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively (Supplemental Table~\ref{table_simulated_model_selection}).
Both the composite and true fitness models outperformed the naive model at 100\% and 89\% of test timepoints, respectively.

@@

Next, we trained and validated models for individual fitness predictors using 25 years of natural A/H3N2 populations spanning from October 1, 1990 to October 1, 2015.
We held out samples collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Figure~\ref{sup_fig:cross_validation_for_natural_populations}).
In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
As antigenic drift is generally considered to be the primary evolutionary pressure on natural A/H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI cross-immunity would be individually more predictive than non-epitope mutations or DMS mutational effects.
Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.

@@

Of the two metrics for functional constraint, non-epitope mutations outperformed DMS mutational effects, with an average distance to the future of 6.14 $\pm$ 1.37 AAs compared to 6.75 $\pm$ 1.95 AAs, respectively.
In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the non-epitope mutations metric was fixed at -0.5, our model learned a consistently stronger coefficient of -0.99 $\pm$ 0.30.
Notably, the best performance of the DMS mutational effects model was forecasting from April 2007 to April 2008 when the major clade containing A/Perth/16/2009 was first emerging.
This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments.
Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive model based on measured mutational effects of a single strain.

LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied HI cross-immunity and non-epitope mutations for the most timepoints outperforming the naive model of 17 (74\%) (Table~\ref{table_natural_model_selection}).
Delta frequency performed worse than LBI and HI cross-immunity and only performed slightly better than non-epitope mutations.
While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.

@@

To test whether composite models could outperform individual fitness metrics for natural populations, we fit models based on combinations of best individual metrics representing antigenic drift, functional constraint, and clade growth.
Specifically, we fit models based on HI cross-immunity and non-epitope mutations, LBI and non-epitope mutations, and all three of these metrics together.
We anticipated that if these metrics all represented distinct, mutually beneficial components of viral fitness, these composite models should perform better than individual models with consistent coefficients for each metric.

@@

Both two-metric composite models performed better than their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
Although the three-metric composite model performed slightly worse than the individual LBI model, this composite model outperformed the naive model the most of any model with 19 (83\%) times out of 23.
The composite of LBI and non-epitope mutations performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs.

@@

As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
The composite model of non-epitope mutations and LBI effectively captured clade growth and decline with 87\% and 89\% accuracy, respectively (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
The estimated best strain from this model was in the top 11th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 78\% of timepoints (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}B).

\begin{figure*}[ht]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.png}
  \caption{
  Validation of best model for natural populations of A/H3N2 viruses, the composite model of non-epitope mutations and LBI.
  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
  }
  \label{fig:validation_of_best_model_for_natural_populations}
  \end{center}
\end{figure*}

@@

After identifying the best model to forecast natural populations, we integrated our forecasting model into our existing open source pathogen surveillance framework, Nextstrain \cite{Hadfield2018}, and used our model to forecast A/H3N2 populations in September 2020.
Our combined non-epitope mutations and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
To aid with identification of potential vaccine candidates, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).

@@

We defined the fitness of each sample at time $t$ as the additive combination of one or more fitness metrics, $f_{i,m}$, scaled by fitness coefficients, $\beta_{m}$.
For example, Equation~\ref{equation_fitness_estimation} estimates fitness per sample by non-epitope mutations ($\mathrm{ne}$) and local branching index ($\mathrm{lbi}$).

\begin{equation}
    f_{i}(t) = \beta_{\mathrm{ne}}f_{i, \mathrm{ne}}(t) + \beta_{\mathrm{lbi}}f_{i, \mathrm{lbi}}(t)
    \label{equation_fitness_estimation}
\end{equation}

>>>>

We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and mutational load (non-epitope mutations) as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
All of these metrics except for HI cross-immunity and DMS mutational effects rely only on HA sequences.
The cross-immunity metrics estimate how antigenically distinct each sample at time $t$ is from previously circulating samples based on either genetic distance at epitope sites or $\log_{2}$ titer distance from HI measurements.
Increased antigenic drift relative to previously circulating samples is expected to correspond to increased viral fitness.
Mutational load is the simplest of the functional constraint metrics, measuring the number of putatively deleterious mutations that have accumulated in each sample since their ancestor in the previous season.
DMS mutational effects provide a more comprehensive biophysical model of functional constraint by measuring the beneficial or deleterious effect of each possible single amino acid mutation in HA from the background of a previous vaccine strain, A/Perth/16/2009.
The growth metrics estimate how successful populations of samples have been in the last six months based on either rapid branching in the phylogeny (LBI) or the change in clade frequencies over time (delta frequency).

@@

We hypothesized that fitness metrics associated with viral success such as epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while metrics associated with fitness penalties, like mutational load, would receive negative coefficients.
We reasoned that both LBI and delta frequency would individually outperform the mechanistic metrics as both of these growth metrics estimate recent clade success regardless of the mechanistic basis for that success.
Correspondingly, we expected that a composite model of epitope cross-immunity and mutational load would perform as well as or better than the growth metrics, as this model would include both primary fitness constraints acting on our simulated populations.

@@

As expected, the true fitness model outperformed all other models, estimating a future population within 6.82 $\pm$ 1.52 amino acids (AAs) of the observed future and surpassing the naive model in 32 of 33 (97\%) timepoints (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Table~\ref{table_simulated_model_selection}).
Importantly, the true fitness model accurately approximated the diversity of future populations coming within -0.29 $\pm$ 2.01 AAs of observed future diversity.
In contrast to the true fitness model, the naive model was an average of 8.97 $\pm$ 1.35 AAs from the future and an average of 2.26 $\pm$ 1.75 AAs from the future's observed diversity.
With the exception of epitope cross-immunity, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
LBI was the best of these models, with a distance to the future of 7.57 $\pm$ 1.85 AAs.
Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
The mutational load metric received a consistently negative coefficient with an average distance of 8.27 $\pm$ 1.35 AAs.

@@

Surprisingly, the composite model of epitope cross-immunity and mutational load did not perform better than the individual mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric when effects of purifying selection were also represented in the model.
The cross-immunity fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \cite{Luksza:2014hj} that are not explicitly present in our simulations.
To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each sample's ancestor in the previous season.
This linear fitness metric did not outperform the cross-immunity metric (Table~\ref{table_simulated_model_selection}).
However, when we fit a composite model of the epitope ancestor and mutational load metrics, this model outperformed both the individual mutational load model and the composite cross-immunity and mutational load model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.

We hypothesized that a composite model of mutually beneficial metrics could better approximate the true fitness of simulated viruses than models based on individual metrics.
To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: mutational load and LBI.
This composite model outperformed all individual metrics with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
The coefficients for mutational load and LBI remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
These results support our hypothesis that multiple complementary metrics can produce more accurate models.

@@

Finally, we tested all of our models on out-of-sample data.
Specifically, we fixed the coefficients of each model to the average values across the training/validation period and applied the resulting models to the next 9 years of previously unobserved simulated data.
A standard expectation from machine learning is that models will perform worse on test data than on validation data.
Instead, we found that nearly all models except for the true fitness and delta frequency performed as well or better with out-of-sample data (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
The composite model of mutational load and LBI outperformed the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively (Supplemental Table~\ref{table_simulated_model_selection}).
Both the composite and true fitness models outperformed the naive model at 100\% and 89\% of test timepoints, respectively.

@@

Next, we trained and validated models for individual fitness predictors using 25 years of natural A/H3N2 populations spanning from October 1, 1990 to October 1, 2015.
We held out samples collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Figure~\ref{sup_fig:cross_validation_for_natural_populations}).
In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
As antigenic drift is generally considered to be the primary evolutionary pressure on natural A/H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI cross-immunity would be individually more predictive than mutational load or DMS mutational effects.
Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.

@@

Of the two metrics for functional constraint, mutational load outperformed DMS mutational effects, with an average distance to the future of 6.14 $\pm$ 1.37 AAs compared to 6.75 $\pm$ 1.95 AAs, respectively.
In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the mutational load metric was fixed at -0.5, our model learned a consistently stronger coefficient of -0.99 $\pm$ 0.30.
Notably, the best performance of the DMS mutational effects model was forecasting from April 2007 to April 2008 when the major clade containing A/Perth/16/2009 was first emerging.
This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments.
Alternate implementations of less background-dependent DMS metrics never performed better than the mutational load metric (Supplemental Table~\ref{sup_table:complete_natural_model_selection}).
Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive model based on measured mutational effects of a single strain.

LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}) and tied HI cross-immunity and mutational load for the most timepoints outperforming the naive model of 17 (74\%) (Table~\ref{table_natural_model_selection}).
Delta frequency performed worse than LBI and HI cross-immunity and only performed slightly better than mutational load.
While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.

@@

To test whether composite models could outperform individual fitness metrics for natural populations, we fit models based on combinations of best individual metrics representing antigenic drift, functional constraint, and clade growth.
Specifically, we fit models based on HI cross-immunity and mutational load, LBI and mutational load, and all three of these metrics together.
We anticipated that if these metrics all represented distinct, mutually beneficial components of viral fitness, these composite models should perform better than individual models with consistent coefficients for each metric.

@@

Both two-metric composite models performed better than their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
Although the three-metric composite model performed slightly worse than the individual LBI model, this composite model outperformed the naive model the most of any model with 19 (83\%) times out of 23.
The composite of mutational load and LBI performed the best overall with an average distance to the future of 5.44 $\pm$ 1.80 AAs.

@@

As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
The composite model of mutational load and LBI effectively captured clade growth and decline with 87\% and 89\% accuracy, respectively (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).
The estimated best strain from this model was in the top 11th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 78\% of timepoints (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}B).

\begin{figure*}[ht]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.png}
  \caption{
  Validation of best model for natural populations of A/H3N2 viruses, the composite model of mutational load and LBI.
  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
  }
  \label{fig:validation_of_best_model_for_natural_populations}
  \end{center}
\end{figure*}

@@

After identifying the best model to forecast natural populations, we integrated our forecasting model into our existing open source pathogen surveillance framework, Nextstrain \cite{Hadfield2018}, and used our model to forecast A/H3N2 populations in September 2020.
Our combined mutational load and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
To aid with identification of potential vaccine candidates, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).

@@

We defined the fitness of each sample at time $t$ as the additive combination of one or more fitness metrics, $f_{i,m}$, scaled by fitness coefficients, $\beta_{m}$.
For example, Equation~\ref{equation_fitness_estimation} estimates fitness per sample by mutational load ($\mathrm{ml}$) and local branching index ($\mathrm{lbi}$).

\begin{equation}
    f_{i}(t) = \beta_{\mathrm{ne}}f_{i, \mathrm{ml}}(t) + \beta_{\mathrm{lbi}}f_{i, \mathrm{lbi}}(t)
    \label{equation_fitness_estimation}
\end{equation}
